---
title: "COVID-19 Vaccines and Economic Recovery in Ecuador 2020-2021."
author: "Daniel Sánchez"
date: "April 2022"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---
# Introduction

This is the preliminary study of COVID-19 Vaccines and Economic Recovery in Ecuador. Through empirical models, I aim to determine if there is any sign of a causal effect of vaccination in economic recovery. I will study this phenonemon from 2020 to 2021 in Ecuador. I try to implement a difference-in-differences design using several variables I created as treatments. 

# Data wrangling
Load all preliminary stuff:
```{r setup}
# Load libraries
library(tidyverse)
library(sandwich)
library(lmtest)
library(modelsummary)
library(fwildclusterboot)
library(fixest)
library(plm)
library(estimatr)
library(psych)
library(openxlsx)
library(corrplot)
library(stats)
library(timetk)
library(xts)
library(lubridate)
library(dygraphs)
library(htmlwidgets)
library(fastDummies)
# Load the data
df<-read.csv('../data/clean_data.csv')
```

I need to wrangle the data before I can start running models.

```{r data-wrangle}
# Create province factors (for the clustered std. errors)

df$prov<-as.factor(df$province)

# Create post dummy variable 

df$post<-ifelse(df$post == 'post',1,0)

# Create the antivax factor variable

df$antivax<-as.factor(df$antivax)

# Relevel

df$antivax<-relevel(df$antivax, 'provax')

# The same for the treatment variable (is exactly equal to the antivax variable, but treat sounds more pro)

df$treat <- as.factor(df$treat)

df$treat <- relevel (df$treat, 'provax')

# Convert the month factor to a factor

df$month<-as.factor(df$month)

df$month<-relevel(df$month, 'January')

# Create a year factor

df$year<-as.factor(df$year)

# Conservate a year numeric variable

df$year_n<-as.numeric(df$year)

# Change the reference level to PICHINCHA

df$prov<-relevel(df$prov, 'PICHINCHA')

```

# Exploratory analysis

## Graphs

### Time series

Here I will plot variables against time to see their evolution. First I need to create a month-year progression (time-series dataframe). 

I first need to wrangle the dataframe a bit and sum by month and year

```{r df-buss-creation}
# Load the data

df_buss<-read.csv('../data/companies.csv')

# Define the date variable and convert to time series object

df_buss$date <- 0 

df_buss$date<-as.Date(df_buss$FECHA_CONSTITUCION,
                      format = "%d/%m/%Y")

df_buss_t<-tk_xts(df_buss, date_var = 'date')

# Create a business creation dummy to sum later. 

df_buss_t$creation<-1

# Now, the actual summ stuff. 

df_buss_monthly<-apply.monthly(df_buss_t$creation,
                               sum)

time_series_buss<-dygraph(df_buss_monthly,
                          xlab = 'Time',
                          ylab = 'Company Creation') %>%
                  dyRangeSelector() %>% 
                  dyHighlight(highlightCircleSize = 5) %>% 
                  dyOptions(axisLineWidth = 1.5,
                            fillGraph = T,
                            drawGrid = F)

saveWidget(time_series_buss,
           file = '../figures/time_series_buss.html')
  
```
### Correlograms
%>% 
I want to plot a simple correlation matrix for my variables, see what's what. 

```{r corplot}
# First create a subset of the data so that the correlation matrix makes sense

df_cor<- df %>% 
         select(buss, vax_leastone, vax_complete, deaths, excessd, excessd_rate, total_cases, case_rate, new_cases,
                new_case_rate, taxes, taxpayers, sales, month_number, transit_acc, v_deaths, year_n, 
                robbery, tax_pc, sales_pc_k, job_change, m_retail, m_grocery, m_parks, m_transit, m_workplace, 
                m_residential)

cor_table<-cor(df_cor,
               use = 'complete.obs') 

corrplot(cor_table,
         method = 'square',
         title = 'Correlogram')


```
Now I'd like to separate the data by year, then draw the same corrplots to see if there is any difference.

For 2020:

```{r corplot-2020}

df_20<-subset(df, df$year == 2020)

df_20_cor <- df_20 %>% 
             select(buss, deaths, excessd, excessd_rate, total_cases, case_rate, new_cases,
                new_case_rate, taxes, taxpayers, sales, month_number, transit_acc, v_deaths,
                robbery, tax_pc, sales_pc_k, job_change, m_retail, m_grocery, m_parks, m_transit, m_workplace, 
                m_residential)

cor_table_20<-cor(df_20_cor,
               use = 'complete.obs') 

corrplot(cor_table_20,
         method = 'square',
         title = 'Correlogram')

```
For 2021:
```{r corplot-2021}

df_21<-subset(df, df$year == 2021)

df_21_cor <- df_21 %>% 
             select(buss, deaths, vax_leastone, vax_complete, excessd, excessd_rate, total_cases, case_rate, new_cases,
                new_case_rate, taxes, taxpayers, sales, month_number, transit_acc, v_deaths,
                robbery, tax_pc, sales_pc_k, job_change, m_retail, m_grocery, m_parks, m_transit, m_workplace, 
                m_residential)

cor_table_21<-cor(df_21_cor,
               use = 'complete.obs') 

corrplot(cor_table_21,
         method = 'square',
         title = 'Correlogram')

```
# Fixed FX Models with vaccination rate treatment

For this section, I try to establish a differences in differences design using the vaccination rate as my treatment variable. 

Run a model which includes the most basic variables. Also run the bootstrapping.

```{r easy-model}
# Model
easy<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete | prov + month,
            cluster = ~ prov,
            data = df)

# Bootstrapping test
easy_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + month + prov, data = df),
                    clustid = 'prov',
                    param = 'post:vax_leastone',
                    B = 9999)
summary(easy_boot)
```
Add the google workplace mobility trend
```{r easy-gog}
easy_g<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace| prov + month,
            cluster = ~ prov,
            data = df)

summary(easy_g)
# Bootstrapping test
easyg_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + month + prov + 
                       m_workplace, data = df),
                    clustid = 'prov',
                    param = 'post:vax_leastone',
                    B = 9999)
summary(easyg_boot)

```

Now consider a model with some more variables. 

```{r complex-g}
complex_g<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                 transit_acc + job_change + tax_pc + sales_pc_k| prov + month,
                 cluster = ~ prov,
                 data = df)

summary(complex_g)

complexg_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                           transit_acc + job_change + tax_pc + sales_pc_k + prov + month,
                           data = df),
                        clustid = 'prov',
                        param = 'post:vax_leastone',
                        B = 9999)
summary(complexg_boot)

```

Add the contagion data to see what happens

```{r complex-contagion}
complex_g_cases<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                       transit_acc + job_change + tax_pc + sales_pc_k + total_cases| prov + month,
                 cluster = ~ prov,
                 data = df)

summary(complex_g_cases)

complexgcase_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + 
                               + robbery + transit_acc + job_change + tax_pc + sales_pc_k + prov + month,
                               data = df),
                               clustid = 'prov',
                            param = 'post:vax_leastone',
                            B = 9999)
summary(complexg_boot)

```

Replaces the total cases with new cases

```{r complex-contagionsnew}

complex_newc<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                       transit_acc + job_change + tax_pc + sales_pc_k + new_cases| prov + month,
                 cluster = ~ prov,
                 data = df)

summary(complex_newc)

complex_newc_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + 
                               + robbery + transit_acc + job_change + tax_pc + sales_pc_k + new_cases + prov + month,
                               data = df),
                            clustid = 'prov',
                            param = 'post:vax_leastone',
                            B = 9999)
summary(complex_newc_boot)

```

Replace with total case rates

```{r complex-contagionrate}

complex_c1<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                       transit_acc + job_change + tax_pc + sales_pc_k + case_rate| prov + month,
                 cluster = ~ prov,
                 data = df)

summary(complex_c1)

complex_c1_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + 
                               + robbery + transit_acc + job_change + tax_pc + sales_pc_k + case_rate + prov + month,
                               data = df),
                            clustid = 'prov',
                            param = 'post:vax_leastone',
                            B = 9999)
summary(complex_c1_boot)

```

Replace with new case rates.

```{r complex-newcaserate}

complex_c1_new<-feols(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + robbery + 
                       transit_acc + job_change + tax_pc + sales_pc_k + new_case_rate| prov + month,
                 cluster = ~ prov,
                 data = df)

summary(complex_c1_new)

complex_c1new_boot<-boottest(lm(buss ~ post*vax_leastone + deaths + year + v_deaths + vax_complete + m_workplace + 
                               + robbery + transit_acc + job_change + tax_pc + sales_pc_k + new_case_rate + prov + month,
                               data = df),
                            clustid = 'prov',
                            param = 'post:vax_leastone',
                            B = 9999)
summary(complex_c1new_boot)

```

Show a table with all of the models that were estimated:

```{r model-table}
modelsummary(list(easy, easy_g, complex_g, complex_g_cases, complex_newc, complex_c1,complex_c1_new),
             stars = c('*' = .1, '**'=0.05, '***' = .01))
```
Now a table with all of the bootstrap stuff:

```{r bootstrap-table}

modelsummary(list(easy_boot, easyg_boot, complexg_boot, complexgcase_boot, complex_newc_boot, 
                  complex_c1_boot, complex_c1new_boot), 
             estimate = "{estimate} ({p.value})", 
             statistic = "[{conf.low}, {conf.high}]")
```

# Fixed FX Models with antivax-provax dummy as treatment and ln(1+x) bussiness creation

Here, I use an antivax-provax dummy as my treatment variable. I'd like to see if provinces which I've labelled as "anti-vax" have a tendency to create lesser businesses. I've based this indicator in the number of cases per capita reported for sicknesses that could have been otherwise prevented by a vaccine (Hep-B, parotiditis, etc.). I use TWFE for this. 

## Data Wrangling

I will change the way I use my dependent variable and apply a ln(1+x) transformation. Also, in order to not have a collinearity problem, I can't estimate the treatment effect, as it is a combination of several province dummies. What I will do is simply include the period and treatment interaction. 

```{r wrangling_alt}
# Create the log transformed variable

df$lbuss<-log(df$buss+1) 

# Also for the non_sas businesses

df$lnsas<-log(df$non_sas+1)

# Create the interaction

df$did<-df$post* df$antivax_dic

```

## Models 

Now I estimate a very simple model just considering the DiD estimator and corresponding FX dummies. Clustering is done by province and months now. I do the wild bootstrap test on the DiD estimator as well. 

First, a simple model simply considering the DiD estimator, the period dummy and the corresponding fixed effects (not shown). Also, compare it with one that uses the interactions between year and month. 

```{r twfe-simple}
simple_twfe<-feols(lbuss ~ post + did | prov + month,
                   data = df,
                   vcov = ~ prov + month)

boot_twfe<-boottest(simple_twfe,
                    clustid= c('prov','month') ,
                    param = 'did',
                    B = 9999,
                    seed = 1) 

simple_twfe_int<-feols(lbuss ~ post + did | prov^month,
                       data = df,
                       vcov = ~ prov + month)

simple_twfe_int_lm<-lm(lbuss ~ post + did + prov*month, 
                       data =df)

boot_twfe_int<-boottest(simple_twfe_int_lm,
                    clustid= c('prov','month') ,
                    param = 'did',
                    B = 9999,
                    seed = 1) 
```

A more complex one adding the most obvious variables, and an important month times province dummy

```{r twfe-asasked}
asked_twfe<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year | prov^month,
                  cluster = ~ prov + month,
                  data = df)
summary(asked_twfe)

asked_twfe_lm<-lm(lbuss ~ post + did + prov*month + vax_leastone + vax_complete + excessd + year,
                  data = df)

boot_asasked<-boottest(asked_twfe_lm,
                       clustid = c('prov', 'month'),
                       param = 'did',
                       B = 9999,
                       seed = 1)

```
Add the google mobility data to these models. 

```{r twfeg}
# With workplace
asked_twfe_g<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_workplace| prov^month,
                    cluster = ~ prov + month,
                    data = df)
summary(asked_twfe_g)

asked_twfeg_lm<-lm(lbuss ~ post + did + prov*month + vax_leastone + vax_complete + excessd + year + m_workplace,
                    data = df)

boot_g<-boottest(asked_twfeg_lm,
                  clustid = c('prov', 'month'),
                  param = 'did',
                  B = 9999,
                  seed = 1)

# With parks

asked_twfe_g1<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_parks| prov^month,
                    cluster = ~ prov + month,
                    data = df)
summary(asked_twfe_g1)

asked_twfeg1_lm<-lm(lbuss ~ post + did + prov*month + vax_leastone + vax_complete + excessd + year + m_parks,
                    data = df)

boot_g1<-boottest(asked_twfeg1_lm,
                  clustid = c('prov', 'month'),
                  param = 'did',
                  B = 9999,
                  seed = 1)

# With residential

asked_twfe_g2<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential| prov^month,
                    cluster = ~ prov + month,
                    data = df)
summary(asked_twfe_g2)

asked_twfeg2_lm<-lm(lbuss ~ post + did + prov*month + vax_leastone + vax_complete + excessd + year + m_residential,
                    data = df)

boot_g2<-boottest(asked_twfeg2_lm,
                  clustid = c('prov', 'month'),
                  param = 'did',
                  B = 9999,
                  seed = 1)

# With all of them mixed

asked_twfe_g3<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace| prov^month,
                    cluster = ~ prov + month,
                    data = df)

summary(asked_twfe_g3)

asked_twfe_g3_lm<-lm(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace +prov*month,
                     data = df)

boot_g3<-boottest(asked_twfe_g3_lm,
                  clustid = c('prov','month'),
                  param = 'did',
                  B = 9999,
                  seed = 1)


```
Add some more variables to a few more models

```{r complex_alts}

asked_twfe_g4<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace + case_rate| prov^month,
                    cluster = ~ prov + month,
                    data = df)

summary(asked_twfe_g4)

asked_twfe_g4_lm<-lm(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace + case_rate + prov*month,
                     data = df)


boot_g4<-boottest(asked_twfe_g4_lm,
                       clustid = c('prov','month'),
                       param = 'did',
                       B = 9999,
                       seed = 1)

asked_twfe_g5<-feols(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace + new_case_rate| prov^month,
                    cluster = ~ prov + month,
                    data = df)

summary(asked_twfe_g5)

asked_twfe_g5_lm<-lm(lbuss ~ post + did + vax_leastone + vax_complete + excessd + year + m_residential + m_parks
                     + m_workplace + new_case_rate + prov*month,
                     data = df)

boot_g5<-boottest(asked_twfe_g5_lm,
                       clustid = c('prov','month'),
                       param = 'did',
                       B = 9999,
                       seed = 1)

```

Now compare all of them

```{r table_google}
modelsummary(list(simple_twfe,
                  simple_twfe_int,
                  asked_twfe,
                  asked_twfe_g,
                  asked_twfe_g1,
                  asked_twfe_g2,
                  asked_twfe_g3,
                  asked_twfe_g4,
                  asked_twfe_g5), 
             stars = c('*' = .1, '**'=0.05, '***' = .01))

modelsummary(list(boot_twfe,
                  boot_twfe_int,
                  boot_asasked,
                  boot_g,
                  boot_g1,
                  boot_g2,
                  boot_g3,
                  boot_g4,
                  boot_g5),
             estimate = "{estimate} ({p.value})", 
             statistic = "[{conf.low}, {conf.high}]")

```
# TWFE with antivax-provax and non-SAS business creation

I change the dependent variable to businesses created which are NOT the newly-introduced SAS legal model in Ecuador. 

```{r as_asked_nsas}

asked_nsas<-feols(lnsas ~ post + did + vax_leastone + excessd + year| prov^month,
                  cluster = ~ prov + month,
                  data = df)

summary(asked_nsas)

# As in the effect, with period and province FE, but without any interaction
asked_nsas1<-feols(lnsas ~ did + vax_leastone + excessd + robbery + v_deaths| prov + monthyear,
                   cluster = ~ prov + monthyear,
                   data = df)

summary(asked_nsas1)


asked_nsasg<-feols(lnsas ~ post + did + vax_leastone + excessd + year + m_workplace| prov^month,
                   cluster = ~ prov + month,
                   data = df)

summary(asked_nsasg)

```
# Justifying paralell trends

I need to try to justify parallel trends for all this to be worth it. I will run a model with period and treatment interactions. First, however, I need to actually create a period dummy. 

```{r data-ptrends}
# Create a year and month variable which I can use for graphs later

df<-
  
df %>% 
mutate(
day = "01", #Create a day column just to get a full date format. The day will be dropped in the following step
date_yr_m = as.yearmon(paste0(year,month,day), "%Y %b") #use the zoo as.yearmon() to get the year and month only
)

# Create short month and year variables

df$month_short<-substr(df$month,1,3)

df$year_short<-ifelse(df$year == 2020, 20, 21)

# Create a factor with basis on month and year

df$my<-paste(df$month_short, df$year_short, sep = '') %>% as.factor()

# Create period dummies

df<-dummy_cols(df, 
               select_columns = 'my',
               remove_first_dummy = F)

# Create the interactions

df$t0120<-df$antivax_dic*df$my_Jan20
df$t0220<-df$antivax_dic*df$my_Feb20
df$t0320<-df$antivax_dic*df$my_Mar20
df$t0420<-df$antivax_dic*df$my_Apr20
df$t0520<-df$antivax_dic*df$my_May20
df$t0620<-df$antivax_dic*df$my_Jun20
df$t0720<-df$antivax_dic*df$my_Jul20
df$t0820<-df$antivax_dic*df$my_Aug20
df$t0920<-df$antivax_dic*df$my_Sep20
df$t1020<-df$antivax_dic*df$my_Oct20
df$t1120<-df$antivax_dic*df$my_Nov20
df$t1220<-df$antivax_dic*df$my_Dec20
df$t0121<-df$antivax_dic*df$my_Jan21
df$t0221<-df$antivax_dic*df$my_Feb21
df$t0321<-df$antivax_dic*df$my_Mar21
df$t0421<-df$antivax_dic*df$my_Apr21
df$t0521<-df$antivax_dic*df$my_May21
df$t0621<-df$antivax_dic*df$my_Jun21
df$t0721<-df$antivax_dic*df$my_Jul21
df$t0821<-df$antivax_dic*df$my_Aug21
df$t0921<-df$antivax_dic*df$my_Sep21
df$t1021<-df$antivax_dic*df$my_Oct21
df$t1121<-df$antivax_dic*df$my_Nov21
df$t1221<-df$antivax_dic*df$my_Dec21

# Add a sum function for it to be more quickly

sum_int_ptrends<-"df$t0120+df$t0220+df$t0320+df$t0420+df$t0520+df$t0620+df$t0720+df$t0820+df$t0920+df$t1020+df$t1120+df$t1220+df$t0121+df$t0221+df$t0321+df$t0421+df$t0521+df$t0621+df$t0721+df$t0821+df$t0921+df$t1021+df$t1121+df$t1221"


```

With this, I can run the regression of the `asasked` model with the period and treatment dummy. 

```{r ptrends-reg}

ptrends_reg<-feols(lbuss ~  vax_leastone + excessd + year + 
                   t0120 + t0220+t0320+t0420+t0520+t0620+t0720+t0820+t0920+t1020+t1120+t1220+t0121+t0221+t0321+t0421+
                   t0521+t0621+t0721+t0821+t0921+t1021+t1121+t1221| prov^month,
                   cluster = ~ prov,
                   data = df)
summary(ptrends_reg)

# Now simply use 2021

df_21<-subset(df, df$year == 2021)

ptrends_reg1<-feols(lbuss ~ post + did + vax_leastone + excessd + t0321+ t0421
                    + t0521+t0621+t0721+t0821+t0921+t1021+t1121+t1221| prov, 
                   cluster = ~ prov,
                   data = df_21)
summary(ptrends_reg1)

# Now eliminate all things but the period dummies and the treatment group, prior to treatment

```
Try simply doing the model to justify parallel trends as seen in The Effect. 

```{r ptrends-time}
# Eliminate all periods before the treatment time (all things prior to September 2021)

df_ptrends<-subset(df, df$time_cont < 21)

# Run the regression

ptrends_contime<-feols(lbuss ~ time_cont*treat,
                       data = df_ptrends)

summary(ptrends_contime, vcov = 'hetero')

# Try with clusters with province

ptrends_contime1<-feols(lbuss ~ time_cont*treat,
                       cluster = ~ province,
                       data = df_ptrends)

summary(ptrends_contime1)

# Try with non-sas and province clusters

ptrends_nsas<-feols(lnsas ~ time_cont*treat,
                    cluster = ~ province,
                    data = df_ptrends)

summary(ptrends_nsas)

# Non-sas and month clusters

ptrends_nsas1<-feols(lnsas ~ time_cont*treat,
                     cluster = ~ month,
                     data = df_ptrends)

summary(ptrends_nsas1)

# Add a quadratic term

ptrends_contime2<-lm(lbuss ~ I(time_cont^2)+ time_cont*treat,
                     data = df_ptrends)

summary(ptrends_contime2)

# Add cubic term

ptrends_contime3<-lm(lbuss ~ I(time_cont^3)+ time_cont*treat,
                     data = df_ptrends)

summary(ptrends_contime3)

# Consider time dummies instead of a continous variable

ptrends_timed<-lm(lbuss ~ my*treat,
                  data = df_ptrends)

summary(ptrends_timed)

# Use clustered errors here, with fixed effects

ptrends_timed1<-feols(lbuss ~ my*treat | prov + my,
                     data = df_ptrends)

summary(ptrends_timed1)

# Use only the province clustered errors, nothing more

ptrends_timed2<-feols(lbuss ~ my*treat,
                      cluster = ~ prov,
                      data = df_ptrends)

summary(ptrends_timed2)

# Same but with the non-SAS

ptrends_timedns<-feols(lnsas ~ my*treat,
                       cluster = ~ prov + month,
                       data = df_ptrends)

summary(ptrends_timedns)

# Use only the month clustered, nothing more

ptrends_timed3<-feols(lbuss ~ my*treat,
                      cluster = ~ month,
                      data = df_ptrends)

summary(ptrends_timed3)

# Only use 2021 and province clusters

df_ptrends21<-subset(df_ptrends, df_ptrends$year == 2021)

ptrends_timed2<-feols(lbuss ~ my*treat,
                      cluster = ~ prov,
                      data = df_ptrends21)

summary(ptrends_timed2)

# Only use 2021 and period clusters

ptrends_timed2<-feols(lbuss ~ my*treat,
                      cluster = ~ month,
                      data = df_ptrends21)

summary(ptrends_timed2)

```
The interaction of the time dummies and/or trend variable is only sometimes non-significant in these test regressions. Notably, when we take out the SAS companies out of the question, there seems to be evidence of parallel trends considering many kind of robust errors. This makes the case for the models which try to explain the creation of non-SAS businesses during the pandemic. Howevers, using month-clustered errors, this seems to go away. So there might a need to include a trend variable in the DiD models. 

Change things a bit and now consider time as a continuous variable. We cluster by province only, and consider time and province interactions. 

```{r reg-time}

# Run the regression

reg_time<-feols(lbuss ~ post + did + vax_leastone + excessd + year  + robbery  + tax_pc +
                sales_pc_k + transit_acc + job_change| prov[time_cont],
                cluster ~ prov + month,
                data = df)

summary(reg_time)

fixef_reg_time<-fixef(reg_time)

# Test the interaction terms

reg_time_pt<-feols(lbuss ~ vax_leastone + excessd + year + robbery  + tax_pc +
                   sales_pc_k + transit_acc + job_change + 
                   t0220 + t0320+t0420+t0520+t0620+t0720+t0820+t0920+t1020+t1120+t1220+t0121+t0221+t0321+t0421+
                   t0521+t0621+t0721+t0821+t0921+t1021+t1121+t1221 | prov[time_cont],
                   cluster ~ prov + month,
                   data = df)

summary(reg_time_pt)
```

Change again things, now eliminating province FE and adding the treatment variable for it to be estimated along with the DID estimator. Use month fixed effects only, although clustering will still be two-way.

```{r reg-noprovfe}
# First, the regression that I'd use for modelling

reg_noprov<-feols(lbuss ~ post*treat + treat*(job_change + robbery + v_deaths + transit_acc + excessd + tax_pc) + 
                  year| month,
                  cluster = ~ prov + month,
                  data = df)

summary(reg_noprov)

# Now, the one with the interactions

# First, relevel the "my" variable for it to be in reference to August, one month before my event

df$my<-relevel(df$my, 'Aug21')

reg_noprovpt<-feols(lbuss ~ treat*my + treat*(job_change + robbery + v_deaths + transit_acc + excessd + tax_pc)| my,
                    cluster = ~ prov + month,
                    data = df)

summary(reg_noprovpt)

# Try restricting to 2021 only

df_21<-subset(df, df$year == 2021)

df_21$my<- as.factor(df_21$monthyear)

df_21$my<-relevel(df_21$my, 'August-2021')

reg_noprovpt21<-feols(lbuss ~ treat*my + treat*(job_change + robbery + v_deaths + transit_acc + excessd + tax_pc)| my,
                      cluster = ~ prov + month,
                      data = df_21)

summary(reg_noprovpt21)



```
# Defining a treatment variable with another way

In this section I'd like to change things a little bit by using a k-means algorithm to define my treatment. I'll use a reduced, "yearly" dataset of provinces with some data on health-related stuff. 

```{r k-means}
# Load the data first
df_prov<-read.xlsx('../data/province_data.xlsx')

# Change the row names so the k-means stuff works
rownames(df_prov)<-df_prov$province

# Delete the province variable

df_prov$province<-NULL

# Normalize the data

df_prov.Z<-scale(df_prov)

# Now do the k-means with 3 clusters and just three variables

df_prov.Z_reduced<-as.data.frame(df_prov.Z) %>% select(diseases_pc, avg_covaxrate, population)

set.seed(1)
kmeans_prov<-kmeans(df_prov.Z_reduced, 3)
print(kmeans_prov)

# Now paste the clusters in the province database

df_prov$cluster <- kmeans_prov$cluster

# Now create a treatment dummy

df_prov$treatment<-ifelse(df_prov$cluster == 1, 1,0)

# Append that information to a new dataframe

# First, return the province names

df_prov$province<-rownames(df_prov)

# Now, create a df with just the treatment variable

df_prov1<-df_prov %>% select(province, treatment)

# Now, the new dataframe

df_new<-df

df_new<-left_join(df_new, df_prov1, by = 'province')

# Create the did variable

df_new$did<-df_new$post*df_new$treatment

# Write a csv file

write.csv(df_new, 
          file ='df_new.csv',
          na = '')

```

We have that some other provinces will fit the treatment as they have relatively low population, high disease rate and relatively low average vaccination rates. 

Update: Stata says this treatment doesn't work (check my [GitHub](https://github.com/dsanchezp18)). 


# Alternative Treatment # 1: Using "rural" provinces

In this section I'd like to try and see what happens if I rethink the treatment and choose treated provinces differently. Here, I consider all of the Amazon region provinces as treated, as well as rural provinces in the Sierra provinces. 

```{r as-asked-treat1}

df$didt1<-df$treat1*df$post

asked_t1<-feols(lbuss ~ post + didt1 + vax_leastone + excessd + year| prov^month,
                cluster = ~ prov + month,
                data = df)

summary(asked_t1)

asked_t1ns<-feols(lnsas ~ post + didt1 + vax_leastone + excessd + year| prov^month,
                cluster = ~ prov + month,
                data = df)

summary(asked_t1ns)


asked_t2<-feols(lbuss ~ post + didt1 + vax_leastone + excessd + year + robbery + v_deaths + transit_acc | prov + month,
                cluster = ~ prov + month,
                data = df)

summary(asked_t2)

asked_t2nsas<-feols(lnsas ~ post + didt1 + vax_leastone + excessd + year + robbery + v_deaths + transit_acc | 
                    prov + month,
                    cluster = ~ prov + month,
                    data = df)

summary(asked_t2nsas)

```
Let us check for parallel trends using prior trends.

```{r ptrends-rural}

ptrends_rur<-feols(lbuss ~ treat1*time_cont,
                   data = df_ptrends)

summary(ptrends_rur, vcov = 'hetero')

ptrends_rur1<-feols(lnsas ~ treat1*time_cont,
                    cluster = ~ prov + month,
                    data = df_ptrends)

summary(ptrends_rur1)

```
No, doesn't really work. 

# Alternative Treatment 3: Indigenous populations

Some Ecuadorian sources have said that those with most vaccine resistance are the indigenous populations. I try switching the treatment to provinces with relatively high indigenous populations. 

```{r as-asked-treat1}

df$didt3<-df$treat3*df$post

asked_t3<-feols(lbuss ~ post + didt3 + vax_leastone + excessd + year| prov^month,
                cluster = ~ prov + month,
                data = df)

summary(asked_t3)

asked_t3ns<-feols(lnsas ~ post + didt3 + vax_leastone + excessd + year| prov^month,
                  cluster = ~ prov + month,
                  data = df)

summary(asked_t3ns)


asked_t3<-feols(lbuss ~ post + didt3 + vax_leastone + excessd + year + robbery + v_deaths + transit_acc | prov + month,
                cluster = ~ prov + month,
                data = df)

summary(asked_t3)

asked_t3nsas<-feols(lnsas ~ post + didt1 + vax_leastone + excessd + year + robbery + v_deaths + transit_acc | 
                    prov + month,
                    cluster = ~ prov + month,
                    data = df)

summary(asked_t3nsas)

```
Let us check for parallel trends using prior trends.

```{r ptrends-rural}

ptrends_ind<-feols(lbuss ~ treat3*time_cont,
                   data = df_ptrends)

summary(ptrends_rur, vcov = 'hetero')

ptrends_ind1<-feols(lnsas ~ treat3*time_cont,
                    cluster = ~ prov + month,
                    data = df_ptrends)

summary(ptrends_ind1)

ptrends_ind2<-feols(lnsas~ treat3*my,
                    cluster = ~ prov,
                    data = df_ptrends)

summary(ptrends_ind2)

```

# Leaving out Pichincha and Guayas from the sample

I'd like to see what happens if I eliminate the two biggest provinces, Pichincha and Guayas.

```{r df_reduced}
df_reduced<-df %>% filter(!province %in% c('PICHINCHA', 'GUAYAS'))

write.csv(df_reduced, file = 'df_reduced.csv', na = "")

```




