---
title: "Immunizing the Economy: A Causality Discussion on Vaccines and Economic Recovery"
subtitle: "Inmunizando a la Economía: Una Discusión de Causalidad Sobre Vacunas y Reactivación Económica"
author: "Daniel Hernan Sánchez Pazmiño"
bibliography: refs.bib
format: 
  html: default
  docx: 
    reference-doc: scvs-template.docx
editor: source
number-sections: true
execute:
  echo: false
  message: false
  warning: false
  cache: true
---

```{r}
# Setup options, for the whole document;
# | label: setup
# | include: false

# Load me my libraries

library(modelsummary) # Regression tables
library(tidyverse) # Tidyverse
library(patchwork) # Multiple plots in one pane
library(fixest) # Fixed Effects OLS
library(openxlsx) # Manage XL stuff
library(fwildclusterboot) # Wild bootstrap tests

# Load me my data

df<-read.csv('../data/df.csv')

```


# Abstract {.unnumbered}

The COVID-19 pandemic severely disrupted economic activity worldwide. In Ecuador, the pandemic found an epicenter in the city of Guayaquil during April 2020, where the excess death rate skyrocketed to almost 700%. However, in mid-2021, Ecuador had a glimpse of hope again as the vaccination process turned extremely effective and under a hundred days half of the country’s population was inoculated. While since then excess deaths have remained at all-time lows in the pandemic era, the Ecuadorian economy remains weakened. This paper discusses a causal effect of vaccination on economic recovery by implementing a two-way fixed effects difference-in-differences estimator on a repeated cross section of Ecuador 2020-2021 at the province-level. Evidence is found that provinces with apparent vaccine-hesitant populations see about 25% less business creation than provinces with openness to vaccination. This finding stresses the importance of fostering vaccination among populations in a country, especially if booster shots are necessary in the future. This effect is found by leaving out the creation of SAS (Sociedad de Acciones Simplificada) businesses, a new firm legal figure introduced to the country in mid-May, which prompted the creation of many new formal businesses due to the reduction of entry costs. Only minimal evidence of differential prior trends is found, which supports the existence of causality. Finally, other estimation methodologies are proposed to identify a more precise causal effect of vaccination on economic activity.

# Resumen {.unnumbered}

La pandemia del COVID-19 interrumpió severamente la actividad económica mundial. En Ecuador, la pandemia encontró un epicentro en la ciudad de Guayaquil en abril del 2020, en donde la tasa de mortalidad en exceso llegó a casi el 700%. Sin embargo, en la mitad del 2021 el Ecuador tuvo un destello de esperanza cuando el proceso de vacunación se tornó extremadamente efectivo y en menos de cien días más de la mitad de población se inmunizó completamente. Aunque desde ese momento las muertes en exceso se mantienen en mínimos históricos en la era de la pandemia, la economía ecuatoriana se mantiene debilitada. Este artículo discute un efecto causal de la vacunación en la reactivación económica utilizando un estimador diferencia-en-diferencias de efectos fijos de doble vía en una sección transversal repetida del Ecuador 2020-2021 al nivel provincial. Se encuentra evidencia de que provincias con poblaciones aparentemente resistentes a la vacunación ven aproximadamente un 25% menos de creación de empresas, en comparación a provincias más abiertas a la vacunación. Este hallazgo resalta la importancia de fomentar la vacunación en las poblaciones de un país, especialmente si dosis de refuerzo se requieren en el futuro. Este efecto se encuentra eliminando la creación de las empresas SAS (Sociedad de Acciones Simplificada), una figura legal de compañía introducida en el país a mediados de 2020 que ocasionó la creación de nuevas firmas debido a la reducción considerable de costos de entrada. Se encuentra evidencia mínima de tendencias previas no similares, lo que apoya a la existencia de causalidad. Finalmente, se proponen otras metodologías de estimación para identificar un efecto causal más preciso de la vacunación en la actividad económica.

# Introduction

Although the 21st century is still far from over, COVID-19 will necessarily be reminded as one of the most important happenings of the time. The changes that the pandemic has brought about have been long-lasting, as the virus which originally emerged in China by late 2019 has had effects still suffered by the time this article was written. In April 2020, the IMF predicted that the economic effects of the pandemic would be the greatest since the Great Depression [@IMF.April2020]. By October 2020, the IMF revised down their projections of world economic growth and the pandemic accumulated over a million deaths to COVID-19, proving to be a worse crisis than what it was initially expected. It was understood that both the health risks posed by COVID-19 and the lockdown policy enacted by the world's governments had greatly hampered economic conditions around the world, especially in the Western Hemisphere [-@IMF.Oct2020].

Ecuador barely escaped economic and social chaos by securing an agreement with the IMF in 2020 worth 6.5 billion dollars after facing terrible consequences due to high public debt and the precipitous fall of oil prices [@IMF.Oct2020.Art]. However, the Ecuadorian economy was still expected to shrink by 11% in 2020 and its unemployment rate was expected to increase in `r 8.1 - 3.8` percentage points relative to 2019 [@IMF.Oct2020]. Guayas became the COVID-19 epicenter in Ecuador with a 900% excess deaths rate [@Naranjo.2020] and with its capital Guayaquil becoming an international attention center due to the extreme humanitarian crisis caused by the COVID-19 impact [@Zibell.2020]. 

However, by 2021 new hope surfaced as newly-developed COVID-19 vaccines received authorities approval and started to be massively distributed around the world by April 2021. The actual economic downturn in 2020 was smaller for both Ecuador and the global economy as well, which encouraged hopes for the coming year. The IMF revised its 2021 projections upward as vaccine rollouts developed in the United States and other developed economies [-@IMF.April2021]. However, on the developing world and notably in Ecuador vaccine distribution became increasingly unequal, as top government officials and private agents received their first vaccine doses before those that actually needed them [@Taj.2021]. In Ecuador the immunization process was initially slow and corrupt: the Ministry of Health's mother and two well-connected TikTok influencers received a vaccination dose before some first-line workers were even eligible to receive one [@Machado.2021; @Sandoval.2021]. Corruption in vaccination and public health contracting played a key role in the resignations and removals of four Ministries of Health between March 2020 and April 2021. 

Yet once again Ecuador became the center of international attention in the second half of 2021 as its sovereign bonds provided the best returns in the world after the possession of economically-liberal President Guillermo Lasso [@Vizcaino.2021]. The Lasso Administration was able to immunize over half of the country's population in 100 days after his possesion in May 24th, which may have converted Lasso into one of the most popular presidents in the region by October 2021 [@Hurtado.2021]. The new Administration's economic agenda promised economic recovery through investment-friendly policy [@Madrid.2021], even marketing the massive vaccination plan as an initial step for recovery after the pandemic shock [@Brik.2021].

The Ecuadorian economy grew 4.2% in 2021 [@BCE.2022], 50% more than the October 2021 IMF projection [-@IMF.Oct2021].The Ecuadorian Central Bank attributes the increases due to important economic gains in the oil, tourism, agriculture and commerce industries, citing the economic effects brought about by the successful vaccine rollout as the main motivator for the internal economic reactivation. However, the statement by the institution may pose an important causal inference issue: even if vaccines have proven to be successful in mitigating COVID-19 infection [@Katella.2022], not much is known about their effect on economic activity. Additionally, the vaccination process may also correlate with other reactivation motivators which may become confounders. Among them stand the `r (78.90/48.52) * 100` % increase in the price of the WTI oil barrel [@MarketWatch.2022] and the loosening of pandemic restrictions which took place after Lasso took office. 

The aim of this paper is to discuss the possibility of a causal effect of COVID-19 vaccination in economic recovery. This is done by implementing a regression-based approach to a difference-in-differences research design to estimate an average treatment effect of vaccine hesitancy in two proxies of economic activity. Province-level data from Ecuador on several economic and demographic indicators published in a monthly basis is used for the econometric analysis, which focuses on the 24 months between January 2020 and December 2021. 

# Literature Review 

In general, literature on COVID-19 and its economic effects which focuses on causal inference is limited. More has been said about causal effects of vaccination and social distancing measures on health outcomes. Causal evidence has been found in support of social distancing and lockdown measures as a means to reduce infections in the COVID-19 pandemic [@Banerjee.2020; @Sharkey.2020] as well as in other pandemics [@Stern.2009; Bootsma.2007; @Mandavilli.2020]. Numerous randomized trials have been run to assess the effects of COVID-19 vaccines, finding that these are safe for mass distribution and effective in slowing down the spread of the virus [@Chen.2021;@Polack.2020;@Yap.2021; @Zheng.2022; @Deb.2021], deeming them essential in the fight against the pandemic.

Regarding economic impacts, it is understood that pandemics hamper economic activity through several channels. According to @Brodeur.2021, these channels are the impact of social distancing measures, direct and indirect costs (hospitalization, loss of labor and production) as well as offsetting and cascading effects as the disruption of services are the most important. The direct channels connect macroeconomically to reductions in household consumption and wealth. The macroeconomic models which were studied by @Brodeur.2021 predict a mostly demand-deficient recession, where policy that benefits the economy's supply may not be successful in inducing recovery. Other studies quantified the opportunity cost of shutdowns in the economy, finding that shutdowns put market production in a 25% production deficit, and that medical development and innovation may reduce the shutdown cost in about 70% [@Mulligan.2021]. Additionally, @Deb.2022 find that containment measures hurt economic activity, associating a containment measure with a 10 percent loss in industrial production over 30 days of measure implementation. Regarding business creation, @Meunier.2022 find that newly registered firms fell in 58% of economies in 2020 relative to 2019, when the normal trend was to see new business creation in 2/3 of economies. 

If COVID-19 was indeed an aggregate demand-intensive shock, it is natural to see a fall in business creation as potential customers become wary of spending in an uncertain environment, where health risks appear to affect the likelihood of spending. If vaccines are indeed the best way to fight the pandemic, it would be expected that economies with higher vaccination rates experience higher economic recovery. The effect of vaccines might be two-fold: on one hand more vaccination may foster less pandemic restrictions which would help economic activity as businesses are less restrained. On the other hand, vaccination may foster economic recovery by reducing the likelihood of contagion and thus increase aggregate demand for goods and services, as the likelihood of contracting COVID-19 is smaller. 

@Furceri.2021 develop an empirical model of vaccination and economic effects across 46 countries, finding a positive statistically significant relationship of vaccination rates and economic activity proxied by nitrogen emissions. The effect could be understood as causal given that "vaccine surprises" are used, the key effect being that an unexpected increase in vaccination per capita is associated with increases in economic activity. Furthermore, it has been said that countries with higher vaccination rates were expected to make stronger economic recoveries according to IMF data, as the projections for regions with higher vaccine inequality were revised downward by April 2021 [@UNDP.2022]. However, this analysis does not make a causal investigation, as indeed it is countries that have been historically richer which had higher vaccination rates in 2022. The lack of causal inference research regarding this key topic is evident, considering that the results of the only study which infers a causal effect may not be applicable to countries outside their sample or at time horizons which can include stronger COVID-19 variants. 

# Methodology

## Data

A province-level database was constructed for this paper. Monthly-published data at the province level was recovered from various government sources to construct a repeated cross-section in which every province has 12 observations on several economic and demographic variables. Thus, the database has `r 24*12` observations.

To proxy economic recovery, data from the Ecuadorian regulatory and supervisory public institution for private companies *Superintendencia de Compañías, Valores y Seguros* (SCVS) is used as a dependent variable. Namely, business creation is used as a proxy for economic recovery, based on the [Companies' Directory database](https://mercadodevalores.supercias.gob.ec/reportes/directorioCompanias.jsf). The number of businesses created in any given month between January 2020 and December 2021 was computed based on the date of creation of the business in the SCVS system. The following table describes all variables used in the paper.

```{r}
# | label: descriptive-table
# Load my data description table from CSv (I did this one on Excel)

data_descrip_df<-read.csv('../data/data_table.csv')

# Print the table using the datasummary command

# Print a pretty table

datasummary_df(data = data_descrip_df,
               output = 'markdown',
               title = 'Variable descriptions and sources')

```
## Empirical approach

A differences-in-differences design is implemented to investigate a potential causal effect of vaccination rates in business creation. The shock that will be exploited for the estimation will be the Delta Variant. This variant of COVID-19 was initially identified in India in late 2020, and spread to the United States in June 2021. It was identified that while the variant's symptoms were the same as the original disease, they seemed to be more severe in the respiratory tract and it also spread quicker. Additionally, the variant seemed to put unvaccinated people at a much higher risk than other variants. It quickly became the principal variant in the United States throughout the fall of 2021 [@Katella.2022; @UCD.2022]. In Ecuador, cases of this variant were initially identified in the summer of 2021 and by October it became the dominant variant in COVID-19 infections [@Mendoza.2021; @Heredia.2021]. This shock is used because given the characteristics of the "new" disease it would be easier to elicit differences in outcomes between groups with different propensity to become vaccinated against COVID-19.

The treatment group will be defined in terms of vaccination hesitancy and actual COVID-19 vaccination rates. In order to accurately exploit vaccination hesitancy and not confound it with logistic/movement restriction issues, the treatment group is defined in reference to the number of vaccine-preventable diseases detected in each province through time. The Ministry of Health has made available yearly vaccine-preventable disease tabloids ([gacetas inmunoprevenibles](https://www.salud.gob.ec/gacetas-inmunoprevenibles/)) which tabulate yearly reported cases of vaccine-preventable diseases at the province level for 2019 and 2020. With this information, the number of vaccine-preventable diseases per thousand is computed with basis to the 2020 population projections (no 2021 projections have been made available). Also, the average monthly vaccination rate is computed, which can be seen at group level in the following table. The treatment group is then composed of the following provinces: Carchi, Chimborazo, Cotopaxi, Morona Santiago, Orellana,  Pastaza, Santo Domingo, Sucumbíos, Zamora Chinchipe. These provinces comply with two conditions: higher than average vaccine-preventable cases per one thousand and lower than average monthly vaccination rates. The following table shows some descriptive statistics for the groups.

```{r}
# | label: group-data-table

group_df<-read.csv('../data/group_stat.csv')

# Print a pretty table

datasummary_df(data = group_df,
               output ='markdown',
               align = 'lllll',
               title = 'Descriptive Statistics for Treatment and Control Groups',
               col.names = c('Group',
                             'New Businesses 2020', 
                             'New Businesses 2021', 
                             'Mean Monthly Vaccination Rate (%)',
                             'Mean vaccine-preventable cases per thousand'))

```
Having defined treatment and control groups and the exogenous shock, the empirical models to be estimated will be based on the following general equation

$$\ln(bc_{it}+1) = \alpha_i + \alpha_t + \beta_1 T_i + \sum_{j = 2}^{m} \left( \beta_j x_{j} \right) + u_{it}$$
where $\alpha_i$ is a set of coefficients for province fixed effects and $\alpha_t$ is a set of coefficients for period fixed effects (24 periods from January 2020 to December 2021). $\hat{\beta_1}$ will be the point estimate for the two-way fixed effects (TWFE) difference-in-differences estimator, as the dummy variable $T_i$ will equal unity for provinces in the treatment group and period (treatment periods are from September to December 2021). Note that $\beta_1$ equals the coefficient of an interaction term between treatment period treatment group dummies, however, the interaction notation is not used here as introducing a treatment group dummy would induce perfect collinearity. In some of the models estimated, the period dummies are dropped and month fixed effects are included instead as a robustness check. Finally, the $\beta_j$ will be the coefficients on time-varying variables which may affect business creation across the groups. All models estimated will be reported with two-way clustering at the province and month level. Wild bootstrap tests [@Roodman.2019] are done on the DiD estimates to verify their significance. 

Two dependent variables are considered. The first one is total business creation, which is equal to the number of businesses created in the SCVS system in any given month and province. The second is non-SAS business creation. This difference is considered because in February 28th, 2020 a new kind of company *Sociedad de Acciones Simplificada* (SAS) was introduced to the Ecuadorian legislation, and as of May of that year it was possible to create this kind of company. This chance in legislation prompted the swift creation of SAS businesses as the process of creation was considerably easier, faster and chepaer than creating the usual PLC and LLC companies [@Gortaire.2020, @Hanze.2020]. It is adequate to believe that this shock may become a confounder in the identification strategy since Ecuador has been determined to be an entrepreneur-propense country in the region but with a considerable number of informal companies and entry costs [@Hora.2021]. Additionally, it has been already determined through the SCVS data that after a year of the introduction of the SAS, there was a considerable increase in business creation, especially in Quito and Guayaquil [@Franco.2021]. 

Given that the treatment variable implies some sort of vaccine hesitancy among the population of the treated provinces, it would be expected that the TWFE DiD estimator is statistically significant and negative if the hypothesis from the Introduction section holds. This is because after a shock in which the pandemic worsens which induces health concerns among consumers, provinces with more vaccine-hesitant populations see less business creation than provinces with less vaccine hesitant populations. The following section reviews the results of the regressions run to estimate the explained models.

# Results

```{r}
# | label: previous-model-data-wrangling
# Some data wrangling for the estimation of the regression models

# Create the dependent variables in log form and relevel factors

df<-
  df %>% 
  mutate(lbuss = log(buss + 1),
         lnsas = log(non_sas + 1),
         treat = as.factor(treat) %>% relevel('provax'),
         monthyear = as.factor(time_cont) %>% relevel('20')) # Period 20 is the reference period for the DiD design

# Create the DiD variable<- the interaction between the period dummy and treatment dummy
# Equivalent to a dummy for observations which are treated in the treatment period
# This is done so we don't fall in a dummy variable trap (Prof. Wooldridge tweeted about this in July 2022)

df$treat_dummy<-ifelse(df$treat == 'antivax', 1, 0)
df$post_dummy<-ifelse(df$post == 'post', 1, 0)

df<-
  df %>% 
  mutate(did = treat_dummy * post_dummy)

```

## Model Results

# References

